---
title: "Diversification of Australian blind snakes"
date: "`r Sys.Date()`"
author: 
output: 
  # bookdown::html_document2:
  #   toc: true
  #   number_sections: FALSE
  bookdown::word_document2:
    reference_docx: [C:/Users/ST/Documents/repo/AusARG_blindsnake/custom-reference-doc.docx]
    number_sections: true
    fig_caption: yes
bibliography: [C:/Users/ST/Documents/repo/anu_thesis_texts/zotero_library.bib]
# biblio-style: "apalike"
csl: jbiogeography.csl
link-citations: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)
# knitr::opts_knit$set(root.dir = normalizePath("..")) #use grandparent folder working directory
# for including graphics we need full path with knitr
WD <- getwd()
```

```{r packages}
library(dplyr); library(readr); library(janitor); 
library(ape); library(phytools);
library(kableExtra); library(RPANDA)
```

**Abstract**

**Taxon**: Blind snakes

**Location**: Australasia

**Methods**:

**Results**:

**Main conclusions**:

**Keywords**
scolecophidia, Diversification, paleoclimate, BioGeoBEARS, GeoHiSSE

\newpage

# Introduction

<!-- General background  -->

The present-day distribution of extant species can provide clues about their biogeographic history and insight into the tempo and mode with which different groups diversified. The aridification of continental Australia began around 20 million years ago, with pockets of arid zone developing and subsequent extreme warming and cooling periods in the last ~5 million years creating what is now known as the 'outback' [@byrneBirth2008]. During this period, arid-adapted vegetation expanded and mesic biomes became more fragmented [@crispHow2013;@martinCenozoic2006]. This period of aridification is believed to have involved the rapid speciation of many arid-adapted vertebrates [@raboskyExceptional2007;@pepperPalaeoclimate2011;@brennanMass2017] and the extinction of some mesic-adapted groups [@bryantLines2016;@cardilloPhylogeny2017;@owenHow2017].

Today, the arid zone covers the majority of area in the Australian landscape (70%) and provides habitat to 43% of all of Australia's terrestrial reptiles, termites, and ants [@byrneBirth2008]. Squamate reptiles (lizards and snakes) represent the largest vertebrate diversity in Australia with over 1000 species (~800 lizards and ~200 snakes), many of which diversified in the last 20 - 6 myr [@brennanMass2017;@oliverPhylogenetic2017]. Examples of reptilest that have colonised Australia from elsewhere and have diversified remarkably from the beginning of the Miocene include agamid dragons [@chenSexual2012], monitor lizards [@brennanPhylogenomics2021], _Sphenomorphus_ and _Eugongylus_ skinks [@skinnerLygosomine2011], _Gehyra_ geckos [@ashmanDiversification2018], elapid snakes [@leeDiversification2016], pythons [@esquerrePhylogenomics2020], and blind snakes [@marinTracing2013]. 

Some clades, like Elapid snakes [@leeDiversification2016;@sandersMolecular2008] and scincid lizards [[@raboskyExceptional2007] showed dramatic ecological diversity taking up different niches. Other clades like geckos and blind snakes remain morphologically conserved and were thought to be species poor until recent molecular data suggested otherwise [@marinHidden2013;@moritzCryptic2018;@zozayaAre2019]. The species richness and distribution of squamate reptiles across different biomes makes them a good group to test hypotheses about the tempo and mode of diversification in Australia. 

Roughly a quarter of the snake species diversity in Australia is represented by blind snakes of the genus _Anilios_, which comprises **47** described species [@uetzQuarter2021]. All currently described species of _Anilios_ are found in Australia except for _A. eryncinus_, which is endemic to New Guinea. Blind snakes are morphologically conserved and often difficult to distinguish, and extensive phylogenetic studies and several efforts to revisit specimens have revealed that the true species diversity is underestimated [@marinHidden2013;@pyronSystematics2014;@hedgesTaxonomic2014;@ellisTyphlopid2019]. The diversification of _Anilios_ in Australia is reported to have originated from a single colonisation event by a mesic adapted ancestor from Asia between 19 - 39 mya [@vidalBlindsnake2010;@marinTracing2013]. However, despite representing a substantial portion of snake biodiversity, there has been few investigations into the macroevolutionary dynamics of blind snakes. The monophyly of the genus makes _Anilios_ well suited to test hypotheses about how lineages diversified with dramatic ecological changes because they are geographically restricted to the continent and can be found in every biome. 

<!-- Environmental correlates of diversification -->
<!-- agamid lizards [@harmonTempo2003], passerine birds [@ricklefsGlobal2006], amphibians [@rollandContribution2019], rosid plants [@sunRecent2020]).  -->
A variety of ecological correlates have been associated with diversification processes of a clade. Among others, studies on the influence of climate (e.g. temperature, precipitation, and latitude) [@ricklefsGlobal2006;@mayhewBiodiversity2012;@condamineMacroevolutionary2013;@cantalapiedraDietary2014;@condamineClimate2019;@rollandContribution2019;@sunRecent2020], elevation [@esquerreHow2019;@boschmanMountain2021], biological interactions [@ramirezAsynchronous2011;@druryEstimating2016;@brennanPhylogenomicsMonitorLizards2021], CO<sub>2</sub> [@kergoatOpposite2018], microhabitats [@bars-closelDiversification2017] and aridification [@raboskyExceptional2007;@pepperGenetic2011;@deepakAridification2018;@catulloAridification2014] have been documented. In Australia, shifts in diversification rates owing to changes in temperature and aridity has been reported in skinks [@raboskyExceptional2007], geckos [@fujitaDiversification2010;@pepperGenetic2011;@pepperPalaeoclimate2011;@oliverIndependent2014;@oliverContrasting2014], frogs [@catulloAridification2014;@vidal-garciaConvergent2015], saltbush [@kadereitMolecular2010], *Acacia* [@millerComparative2013], insects [@owenHow2017;@toussaintUnveiling2015], and other plants (reviewed in @crispHow2013).

There are several links between temperature and diversification rates of reptiles. First, warm temperature may increase resource availability, which in turn increase ecological niche in general [@clarkeClimate2006]. Second, spatial and temporal heterogeneity in temperature may lead to climatic niche divergence, such as local adaptations in different geographic areas [@pepperGeodiversity2013], and change in activity time of prey/predators [@santosTemperature2011;@lawsonLatitudinal2014]. Third, change in climate (e.g. glaciation and aridification) may cause fragmentation of continuous habitats leading to isolated populations, which may lead to divergence [@oliverRocks2019]. Fourth, ectotherms are subjected to higher metabolic rates in warm climates (i.e. Kleiber's law), which may increase mutation rates and ultimately greater speciation rates [@allenKinetic2006;@makarievaMean2008]. Finally, aridification of Australia may offer new ecological niche for species to exploit, which have been shown to result in deep divergences in many arid-adapted [@byrneBirth2008;@pepperGenetic2011;@marinTracing2013]. 

The influence of temperature on diversification rates has been empirically tested with birth-death models [@condamineMacroevolutionary2013;@kergoatOpposite2018;@condamineClimate2019;@rollandContribution2019], these temperature data are reconstructed (interpolated) based on conversion of concentration of oxygen isotopes [@condamineMacroevolutionary2013]. However, the relationship between diversification rates and aridity are attributed to some proxy such as the the timing of aridification or global temperature rather than quantifying the extent of aridity itself [@dornStrange2014;@deepakAridification2018;@catulloAridification2014;@fujitaDiversification2010;@pepperGenetic2011;@pepperPalaeoclimate2011;@oliverIndependent2014;@oliverContrasting2014;@owenHow2017;@crispHow2013;@toussaintUnveiling2015]. As such, no study has been able to quantitatively assess aridity as a continuous variable in environment-dependent models and compare it to other ecological processes. Recent development in geographic-based methods such as `gen3sis` modeling engine [@hagenGen3sis2021] enables us to simulate climatic, geological, and ecological factors that are spatially and temporally specific to certain areas. In practice, these outputs from `gen3sis` provide paleotemperature and aridity data specific to Australia in the time span of interest, between the 30 MYA to present. These datasets will allow us to empirically test the influence of localised ecological processes on diversification rates over time. 

The aim of this study is to infer the ancestral biome of Australian blind snakes based on a near-complete phylogeny of extant taxa. Additionally, we aim to investigate how paleo-environmental changes have influenced the rates of diversification within this group. To do this, we assess the fit of models between paleo-environmental conditions and diversification rates, and compare the rates of diversification between lineages in the arid versus non-arid biome. 

<!--- In this study, we use the Australian radiation of blindsnakes as a system to investigate the diversification of morphologically conserved squamates by testing hypotheses about the processes that promote speciation. Specifically, we sought to answer the following questions:
(1) What is the ancestral biome of Australian _Anilios_? 
() Which biomes are the largest source and sinks of biodiversity? 
() What is the relationship between diversification rate and changes in paleo environment? 
() Which paleo environment best explain current phylogeny?
() Are there differences in diversification rates between lineages in arid and non-arid biome? --->

#  Materials and methods

## Phylogenetic hypothesis and taxon sampling
<!-- Based on our phylogenetic tree -->
We employed a time-calibrated phylogenomic tree of Australian blind snakes with outgroups (Tiatragul et al. in review). This tree includes 38 operational taxonomic units (OTUs) representing 36 described species of _Anilios_. However, our phylogenetic data suggest the existence of some cryptic species, so we adjusted our taxon sampling fraction to 38/50 species. The genetically distinct OTUs that are not fully recognised as species include 3 '_Anilios grypus_' (2 in the west and 1 in the east) and 2 '_Anilios ligatus_' (one from the east and one from the north). For ancestral biome estimation analyses, we also included _Ramphotyphlops multilineatus_, _Sundatyphlops polygrammicus_, and _Acutotyphlops subocularis_ as outgroup taxa.s

## Biogeography and biome transitions

To investigate the biogeographic history of blind snakes, we used the R package `BioGeoBEARS` 1.1.1 [@matzkeBioGeoBEARS2013]. BioGeoBEARS provides a flexible, likelihood-based framework that models how branches in the phylogeny (representing extant taxa and their ancestors) have shifted among a set of discrete biomes over time. These shifts in biomes can occur within a single branch (anagenetic events, including processes such as dispersal, extinction, or switching to adjacent biomes) or at the time of branching (cladogenetic events involving sympatric, vicariant, and founder-event speciation). The fundamental inputs for the software include a time-calibrated phylogeny, information about the biome/range in which each species is found, and a historical biogeographic model. We divided Australo-Papua into nine discrete biomes and assigned each Australian species to one or more biomes based on their occurrence records. Biome classifications were based on the World Wildlife Fund's (WWF) Terrestrial Ecoregions of the World dataset [@olsonTerrestrial2001] and followed the nomenclature of published Australia-specific studies [@cardilloPhylogeny2017;@pavon-vazquezComprehensive2021].

We applied the phylogeny and distribution information to widely used historical biogeographical methods, including likelihood implementations of Dispersal-Vicariance Analysis(DIVALIKE) [@ronquistDispersalVicariance1997], Dispersal-Extinction-Cladogenesis (DEC) [@reeMaximum2008], and BayArea (BAYAREALIKE) [@landisBayesian2013] in `BioGeoBEARS` [@matzkeBioGeoBEARS2013]. We included two additional free parameters that we thought were relevant. The first parameter, *j*, assigns per-event weight to founder-events [@matzkeModel2014], and the second parameter, *x*, multiplies the probability of dispersal by distance to the power of x [@vandamEvaluating2016] (see Appendix-method-biogeobears).
We used likelihood ratio tests to determine if the additional free parameter was a justified improvement over the simpler base models, and selected the model that best fitted the data based on corrected Akaike Information Criterion (AICc) scores. Based on the best model, we ran an additional 50 biogeographic stochastic maps to account for uncertainty in ancestral biome estimation and transition between biomes, and to estimate the frequency and directionality of dispersal events across biomes.[@matzkeBioGeoBEARS2013].

## Paleo environment reconstruction
<!-- Consult Alex Skeels -->
To estimate paleoclimate, we used three paleographic digital elevation models (DEMs) from @scoteseAtlas2021, @straumeGlobal2020, and [VALDES citation]. Each DEM provides paleotopography at a resolution of 1&deg; x 1&deg; and 5 myr resolution. We projected the paleoenvironmental reconstructions into an equal-area Behrmann projection of 110km x 110km, following the methods described in @hagenGen3sis2021. From the projected data, we extracted temperature and aridity values across all grid cells that fall within Sahul by manually drawing a polygon that encompassed the continent from 40 mya to the present day.
Based on the ____ value, we use XX as the threshold for arid. Thus aridity index is ...

<!--- Because reptiles are sensitive to extremely low and high temperatures, we decided to summarise the reconstructed paleo temperature data into three different datasets: minimum temperature, maximum temperature, and mean temperature. --->

## Time- and environment-dependent diversification analyses 

To evaluate how diversification rates correlated with climate and aridity through time, we fitted several time- and environemnt-dependent likelihood diversification models using the `fit_bd` and `fit_env` functions in the R package `RPANDA` [@morlonRPANDA2016] to the _Anilios_ phylogeny. Specifically, we fitted different combinations of pure-birth and birth-death models with speciation and/or extinction rates being constant or otherwise varying linearly or exponentially with paleo environment as described above. We also fit the model to widely use mean global paleo temperature data set [@condamineMacroevolutionary2013]. To account for missing taxa, we used the fractioning function (f = 38/50) in `RPANDA`. We then assessed which diversification model best fit the data based on AICc scores.

## Geographic range and rates of diversification 

To test whether diversification rates differ between lineages in arid biomes and non-arid lienages, we used two approaches. First, we estimated speciation rates at the origin of each branch (hencforth branch-specific rates) and tip speciation rates using the Julia package `ClaDS` [@malietFast2022]. `ClaDS` is designed to capture frequent shifts in speciation rates with small effects. At each speciation event, daughter lineages are assigned its own speciation rate sampled from a distribution parameterised on the parental rate [@malietFast2022]. 

In the second approach, we fit four state-dependent speciation and extinction (SSE) models to the phylogeny and geographic range data under the GeoSSE [@goldbergPhylogenetic2011] and GeoHiSSE [@caetanoHidden2018] framework using the R package `hisse` v. 2.1.9 [@beaulieuDetecting2016]. 
Since these SSE models only allow two states, we restricted our geographic states to arid and non-arid biomes. Based on the same categories we used in our BioGeoBEARS analyses, we classified species that primarily occur in the arid biome as arid (State 0; n = 8), species that occur primarily in other biomes as non-arid (State 1; n = 24), and those that occupy both the arid and non-arid biomes as widespread (State 01; n = 5).

The four SSE models differed in whether they allowed diversification rates to be dependent on geographic states (range-dependent) or independent of them (range-independent), and whether they allowed an unmeasured trait ('hidden states') to influence diversification rates [@beaulieuDetecting2016;@caetanoHidden2018]. 
Model 1 assumed equal diversification rates regardless of geographic state. Model 2 allowed geographic range to affect diversification rates (the canonical geoSSE model).
Models 3 and 4 were based on models 1 and 2, but they also allowed unmeasured traits ('hidden states') to contribute to diversification rates (see Table GeoHiSSE). By incorporating hidden states, we could assess whether geographic range or an unmeasured trait contributed more to the observed diversification. For each model, we assumed uniform taxon sampling for 89% of arid lineages, 71% of non-arid lineages, and 83% of widespread lineages, based on the number of species in the phylogeny and all currently described species of _Anilios_. We then calculated AIC to identify the best-fitting models. Following the approach in @caetanoHidden2018, we averaged the estimates of diversification parameters and ancestral geographic range estimate across all models according to their AIC weights.

## Data visualisation and software
ClaDS was implemented in `Julia` v.1.8.1 [@bezansonJulia2012], all other analyses were performed in `R` v.4.2 [@rcoreteamLanguage2022]. Figures and graphs were prepared using a combination of base R functions and `ggplot2` [@wickhamGgplot22016]. All map graphics were created using the Free and Open Source `QGIS` v.3.6. We arranged and finalised all figures using Adobe Illustrator v2022.

# Results

## Ancestral biome reconstruction and transitions

<!-- ## values and info from 04_BioGeoBEARS_bsm_plots.R -->
Our BioGeoBEARS analyses all inferred tropical grassland as the ancestral biome of *Anilios* but the ancestor leading to the crown *Anilios* is ambiguous. We found that the DEC+_j_+_x_ model was the most preferred biogeographical history model (delta AICc = -4.47, see Table-supp-BioGeoBEARS). A likelihood ratio test (LRT) showed that both founder events (+*j* parameter) and distance between biomes (+x parameter) together provide a statistically significant improvement over the simpler DEC model, DEC+x, and DEC+j (χ^2^, DF = 2, D-statistic = 26.7, p-value < 0.001).

Based on 50 BioGeoBEARS biogeographic stochastic mappings (BSM) of the DEC+_j_+_x_ model, we estimated that an average of 71.2 ± 1.3 (mean ± SD) total anagenetic and cladogenetic events occurred throughout the phylogeny (see Table-supp-BioGeoBEARS). Of these 40 ± 0 were cladogenetic events compared to 31.2 ± 1.3 anagenetic dispersal events. Out of the 40 cladogenetic dispersal events, most were narrow sympatry range expansion events (23.8±2.1 events) followed by subset sympatry (5.2±2.3 events), vicariance (4.9 ± 1.3), and founder events (6.1 ± 1.5). Our model also indicate that range expansion is expected to decrease as distance increase (free parameter _x_ = -1.3). 

The largest source of outward dispersal is the tropical grassland biome (16.2 ± 4.9 events; 43.6%), which makes up almost double the number of outward dispersals from the arid biome (9.7 ± 3.2 events; 26.0%) (\@ref(fig:fig-biogeobears))c-d. Together the tropical grassland biome and the arid biome contributes the largest source of diversification accounting for 69.6% of outward transitions. On the other hand, the largest sink for all inward dispersals are the tropical forest biome (7.9 ± 4.2 events; 21.2%) and the mediterranean biome (7.0 ± 2.9 events; 18.8%). The greatest directional transition is from the arid biome to the mediterranean biome (5.3 ± 1.5 events) and from the tropical grassland into the temperate forest biome (5.1 ± 1.3 events) (\@ref(fig:fig-biogeobears)c-d.

Based on current distributions from ALA, most _Anilios_ species are found entirely or at least partially in the tropical grassland biome (n = 20) followed by the arid biome (n = 13) and mediterranean biome (n = 10). We identified divergent biogeographic transitions between two major groups of species, which we will refer to as *Group 1* (n=21 lineages) and *Group 2* (n=14 lineages) (\@ref(fig:fig-biogeobears)a. The common ancestor of *Group 1* remained mostly in the tropical grassland biome, then 12/21 lineages expanded or transitioned into other biomes. The common ancestor of *Group 2* first transitioned into the arid biome, and then 7/14 lineages expanded to the mediterranean and temperate forest biomes, and/or transitioned back into the arid biome. 

## Paleoclimate estimates

Paleoclimate reconstructions of aridity differed slightly between the two methods \@ref(fig:fig-paleoclimate-compare)a. Both aridity index shows increase in aridity post 10 MYA, consistent with the onset of aridification of the continent [@byrneBirth2008;@pepperLife2021]. Paleoclimate estimates for Australia differed between the three methods (\@ref(fig:fig-paleoclimate-compare))b. However, all three paleo temperature models indicate Australia was hotter than the average global temperature. Paleoclimate in Australia were also relatively constant compared to global temperature until the Plio-Pleistocene glacial cycles. 

## Environment-dependent diversification 

We found that the exponential aridity-dependent speciation and constant extinction (BEnvDCST) model based on DEM from @scoteseAtlas2021 provided the best fit to the phylogeny (AICc = 238.43; see Table-RPANDA-TOP10). This best-fitting model indicates that an increase in aridity is exponentially associated with a decrease in speciation rate (lambda = 1.32, alpha = -0.14) and a constant background extinction rate (mu = 0.23). The second and third most competitive models (delta AICc < 2) also found that an increase in aridity is exponentially associated with a decrease in speciation rate, but without estimating extinction (see Table-RPANDA-TOP10). Among the exponential environment-dependent models, we found that Australia-specific paleoclimate data provided a better fit to our phylogeny than global paleoclimate data (see supp-TABLE_FULL_RPANDA_model_fit).

## Diversification rates between arid and non-arid biomes

We applied ClaDS and SSE models to evaluate whether diversification rates rates differ between lineages in arid, non-arid, and widespread ranges. Results from our ClaDS analysis show mean branch-specific rates did not differ significantly between lineages in arid, non-arid, and widespread ranges (\@ref(fig:fig-clads_sse). ClaDS also indicate branch-specific diversification rates gradually decrease towards the present.

In our SSE analyses, we found that the range-independent model fit the data better than models where diversification rates were dependent on geographic range (see supp-table-geohisse). The best-fitting geographic range-independent model included a hidden unobserved state, indicating that diversification rates are better explained by this hidden state. Compared to the three other models, the best-fitting model accounted for 95% of the AIC weight and had an AIC value 7.02 units lower than that of the second best-fitting model (see supp-table-geohisse). The weighted average parameter estimates from all models indicate that both speciation rates and turnover rates did not differ significantly between arid and non-arid ranges (see \@ref(fig:fig-clads_sse) and supp-fig-clads-sse for net diversification rate and extinction rates).

# Discussion

With 47 recognised species, the mostly Australian-endemic genus *Anilios* is currently the most diverse genus of scolecophidia in the world. In this study we inferred the extant species in Australia has a tropical grassland origin. Within a few million years, this ancestral lineage split with one group staying in the tropical grassland biome while the other group transitioned into the arid biome. Subsequent transitions in these two groups resulted in the distribution of biomes we observed across the phylogeny today. We also found that paleo aridity index provides the best fit to diversification dynamics observed in our blind snake tree. Although aridification has been linked elevated diversification rates in other plants and animals [@brennanMass2017;@raboskyAnalysis2014;@hernandez-hernandezAridification2014;@jhaPhylogeny2021], we found no such association for Australian *Anilios* blind snakes. 

## Ancestral biome of Australian *Anilios* 

Our result supports the hypothesis that the ancestor of the Australian clade first invaded the mesic part of the Australian continent between 25.8 – 16 MYA [@marinTracing2013]. This is also consistent with other radiations of Australian squamates, where the ancestral lineage first inhabited tropical-mesic regions before later expanding into other biomes as they emerge [@chappleParallel2004;@melvilleEvolutionary2011;@oliverPhylogenetic2017]. Given what we know about the current distribution of species included in our phylogeny, our analyses support the idea that *Anilios* have non-Australian origins. However, whether the Australian radiation was a result of an oceanic dispersal from Southeast Asia or Indonesia as proposed by @vidalBlindsnake2010 is equivocal. Instead, our results suggest the ancestor of Australian _Anilios_ most likely occurred either in the Bismarck Archipelago, New Guinea, or a combination of both. Increasing taxonomic sampling, particularly by including the New Guinean endemic _Anilios erycinus_ and other closely related Typhlopidae snakes (e.g. *Ramphotyphlops*, *Acutotyphlops*) as outgroups will enable us to make more informed inferences about the origin prior to colonising Australia. 

## Biome transition and current distributions 

Consistent with previous research by @marinTracing2013, we found that the initial diversification of _Anilios_ occurred in the Tropical Grassland biome, followed by transitions and/or expansion into other biomes. However, our analyses indicated that species-level divergence occurred earlier, between 20 – 10 MYA, compared to the 15 – 10 MYA proposed by @marinTracing2013. Our BioGeoBEARS analyses suggest that _Anilios_ gradually diversified in the tropical grassland (Group 1) and arid biome (Group 2) during a time when Australia was relatively wet and humid [@byrneBirth2008]. This indicates that many species found in arid biomes, particularly in Group 2, diversified in the arid biome prior to the onset of aridification in the late Miocene [@byrneBirth2008]. 

A hypothesis to explain the contemporary distribution of _Anilios_ and other blind snakes in arid biomes is that they may be pre-adapted to living in such environments. Blind snakes have been found in xeric areas around the world, such as _Rena humilis_ in the desert of North America and _Myriopholis macrorhyncha_ in semi-arid areas from east Africa to the Middle East [@sindacoReptiles2013]. To our knowledge, these species share similar dietary and morphological specialisations with their closest relatives, which is a pattern identified across all fossorial snakes [@cyriacDigging2018]. 

## Influence of paleoenvironment on diversification rates 

Our best-fitting model show that aridity is negatively associated with speciation, where speciation decreased as aridity increased. Two of our reconstructed paleo aridity index show that aridity increased starting ~25 MYA ago with particular sharp increase from 10 MYA and intensification from 5 MYA onwards. Out of the 38 lineages we considered, 5/36 divergence event occured in the last 10 million years when the continent intensively aridified. In contrast, 31/36 divergence event occured between 25 - 10 MYA when the continent was getting warming but not entirely arid. Indeed 12/16 lineages that have current distributions in the arid biome diverged before 10 MYA. This pattern is consitent with other studies on vertebrates that have shown that the divergence time for some arid-adpated taxa pre-dated the intense aridifcation in the last 5 million years [@byrneBirth2008;@catulloAridification2014]. 

While not the best fitting model, we observed that our reconstructed Australia-specific paleotemperature data provided a better fit than the global temperature data set [@condamineMacroevolutionary2013]. In the context of our birth-death models, the absolute value of temperature might not have made much of a difference because our environment-dependent models fit the spline rather than absolute temperature. However, models that take in account thermal niche tolerance, such as NEMo [@huaAdapting2022], will benefit from paleoclimate reconstructions such as these to inform limits (boundaries). We suggest future studies that aim to understand how aridification affect diversification rates in Australia consider using these paleoclimate models to inform model inferences. 

ALEX SKEELS add to this discussion?
- Paleoclimate reconstructions based on DEM by Scotese and Straume were similar in profile. However, DEM base on Valdes differed from the other two. Scotese is characterised by repeated fluctuations in the Pleistocene to reflect glacial cycles. 

## Diversification in the arid zone

Several studies have shown that arid-adapted species originated either from a single ancestor present in Australia before the Miocene, or through multiple independent expansion into the arid zone [@byrneBirth2008]. Our ancestral biome analyses show that several non-arid ancestors of _Anilios_ blind snakes have diverged into the arid zone evident by multiple arid species throughout the phylogeny. This pattern is consistent with several other groups of reptiles that started to diversify in the Miocene including goannas [@brennanPhylogenomicsMonitorLizards2021;@pavon-vazquezCompetition2022], elapid snake [@sandersMolecular2008], pythons [@esquerrePhylogenomics2020] and several geckos [@fujitaDiversification2010;@oliverIndependent2014]. 

For these terrestrial reptiles, increased rates of speciation for arid adapted taxa have been attributed to species filling niche in the saxicolous and xeric habitats that emerged with the contraction of mesic forest throughout the Miocene [@sandersMolecular2008;@hugallPhylogeny2008]. However, results from our weighted-averaged state speciation-extinction analyses on the Australian blind snake phylogeny indicated that speciation, extinction, and turnover rates did not differ between lineages in arid and non-arid biomes. One hypothesis is that fossorial species may not be subjected to the same selective pressures present in the arid zone, such as extreme temperatures [@pepperLife2021], as much surface dwelling species do. For instance, because blind snakes live in subterranean microhabitat with relatively stable climate compared to the surface [@kinlawReview1999;@nevoAdaptive1979], modification from the already specialised phenotype may not be selected for [@cyriacDigging2018]. Future studies that investigate phenotypic variations across all blind snakes will further our understanding of how these snakes evolved with seemingly conserved phenotypes.


- Future phylogeographic study should consider densely sampling species to see whether there are more expansion into or out of arid zone for widespread taxa. 

Warmer and dryer environment may not directly influence speciation rate because blind snakes are physically adapting to warmer temperatures. 
However, climatic fluctuations can contract and fragment available niche for blind snakes and thus influence diversification rates. 

- Microhabitat better explains diversification than climate, but typhlopidae has the second highest speciation rates among fossorial snakes [@bars-closelDiversification2017]. 


- Interestingly blind snakes have diversified greatly compared to other fossorial vertebrates. 
  - It is also possible that fossorial species are more cryptic and thus the full extent of their diversity is under appreciated. 

##### Geographical things
- Geographic barrier for mesic adapted lineages in the temperate and tropical forests may have difficulty crossing the lines [@bryantLines2016]
- Certain regions with complex topographic features tend to harbour more diveristy than flat featureless desert areas [@pepperPalaeoclimate2011]. 
- Lineages that have evolved more recently in the arid zone became extinct with the Plio-Pleistocene glacial cycles leaving only species that have survived in pockets of refugia. 

- The early burst of speciation is observed in lineages that colonise new areas. For example ... 
- Extinction is a big cavaet especially because we suspect there are more species. Arid adapted lineages may have experienced higher extinction OR not well sampled in remote Western Australia. Most species are sampled next to roads [CITE Xia's paper or do supplemental analyses].


<!---
@pepperGenetic2011 - mountain ranges harbour more diversity than deserts. arid zone vertebrates show little genetic diff. But recent population expansion in the desert. 
@catulloAridification2014 - multiple independent origin from mesic or monsoonal ancestors into the arid zone. Lineages older than intense aridification. 
@fujitaDiversification2010 - independent origins of arid zone lineages in _Heteronotia_
@oliverIndependent2014 - multiple independent invasion into the arid zone.
@owenHow2017 - cicada increase speciation in arid zone compared to mesic zone (SSE)
@toussaintUnveiling2015 - diving beetle boom and bust diversification. Invaded arid zone and changed ecologically
@deepakAridification2018 - India accelerated speciation with aridification from the late Miocene. Probably due to emergence of open habitats
--->



## Limitations
There are several limitations in our models. 

- Without extensive sampling we can not confidently trace the pathways in which _Anilios_ colonised Australia. Due to the coarseness of our biomes, we also cannot infer exactly where the ancestral species landed in Australia. Current biogeographic models does not account for extinct species and cannot model complex dynamics of geological processes such as continents moving. 

- Limitations is we can only coarsely assign species to one biome one some are range-limited endemics on islands. For example *A. longissimus* has only been collected from 30m underground on Barrow Island located XX km the northwest coast of Western Australia. Placing that species with arid biome is not realistic but that area is considered the arid biome by the WWF Ecoregions map.

- While taxonomic sampling the most representative of the genus, it is not comprehensive. We suspect there remains several undescribed species of blind snakes in Australia and New Guinea that may belong in the genus _Anilios_. Getting genetic data from _A. erycinus_ will allow us to place the species among other _Anilios_ and provide a more comprehensive picture of the evolution of this genus. 

- Our RPANDA model did not model environment-dependent extinction. 

Therefore, we cannot make any conclusions about transitions between New Guinea and Australia even though at least two species (i.e. _A. torresianus_ and _A. leucoproctus_) are found between southern New Guinea and Far North Queensland 

- Need to account for heterogeneity of the arid zone [@pepperLife2021].


# Conclusion

As we continue generate more genetic data to build ever more extensive phylogenies, there is a growing need to develop methods and complimentary paleo environmental datasets that allow us to test hypotheses about how different macroevolutionary processes influence diversification. We show that tools like `gen3sis` can be used to reconstruct localised paleoaridity data that provides a better fit to our Australian blind snake phylogeny than generic global paleotemperature. To our knowledge, this is the first study that empirically fit diversification models to reconstructed paleo-aridity data. We believe these regional paleoclimate models will be of interested for studies where species are geographically restricted. 




# Tables

Table BioGeoBEARS. BioGeoBEARS model estimates and model AICc scores. Lower AICc scores indicate model is preferred. + Log likelihood
Table RPANDA. Diversification model fit estimates and AICc score of model fit

# Figure legends

Fig. Ancestral biome: Stochastic map of Ancestral biome from Australo-Papuan blind snakes
Fig. Paleoclimate: Reconstructed paleoclimate values from different data sets. Mean temperature data. Aridity index based on precipitation threshold. Abbreviations Pli = Paleocene and Ple = Pleistocene. 
Fig. Speciation rates + phylogeny over time + lineages through time

\newpage 

**References**
<!-- # References -->