---
title: "Diversification of Australian blind snakes"
date: "`r Sys.Date()`"
author: 
output: 
  # bookdown::html_document2:
  #   toc: true
  #   number_sections: FALSE
  bookdown::word_document2:
    reference_docx: [C:/Users/ST/Documents/repo/AusARG_blindsnake/custom-reference-doc.docx]
    number_sections: true
    fig_caption: yes
bibliography: [C:/Users/ST/Documents/repo/anu_thesis_texts/zotero_library.bib]
# biblio-style: "apalike"
csl: jbiogeography.csl
link-citations: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)
# knitr::opts_knit$set(root.dir = normalizePath("..")) #use grandparent folder working directory
# for including graphics we need full path with knitr
WD <- getwd()
```

```{r packages}
library(dplyr); library(readr); library(janitor); 
library(ape); library(phytools);
library(kableExtra); library(RPANDA)
```

**Abstract**

**Taxon**: Blind snakes

**Location**: Australasia

**Methods**:

**Results**:

**Main conclusions**:

**Keywords**
scolecophidia, Diversification, paleoclimate, BioGeoBEARS

\newpage

# Introduction

<!-- General background  -->

Present-day distrubtion of extant species offers clues about their biogeographic history and how the lineages may have diversified. 

The aridification on continental Australia over the past 20 million years resulted in a diversity of habitats and spurred multiple radiations of flora and fauna across the continent, many of which are endemic [@byrneBirth2008;@melvilleEvolutionary2011]. The arid zone makes up the majority of the Australian landscape and provides habitat to ~43% of all Australia’s terrestrial squamate reptiles (lizards and snakes). Reptiles represent the largest vertebrate diversity in Australia with over 1000 species (~800 lizards and ~200 snakes), many of which diversified throughout the Miocene (starting ~25 MYA) to the present [@oliverPhylogenetic2017]. For example, agamid dragons [@chenSexual2012], monitor lizards [@brennanPhylogenomics2021], _Sphenomorphus_ and _Eugongylus_ skinks [@skinnerLygosomine2011], _Gehyra_ geckos [@ashmanDiversification2018], elapid snakes [@leeDiversification2016], pythons [@esquerrePhylogenomics2020], and blind snakes [@marinTracing2013] are reptiles that colonised Australia from elsewhere and have diversified remarkably from the beginning of the Miocene. 

Roughly a quarter of the snake species diversity in Australia is represented by blindsnakes of the genus _Anilios_, which comprises **47** described species [@uetzQuarter2021]. All _Anilios_ species are found in Australia except for _A. eryncinus_, which is only found in Papua New Guinea. While morphologically conserved and difficult to distinguished, extensive phylogenetic studies and several efforts to revisit specimens revealed the true species diversity is underestimated [@marinHidden2013;@pyronSystematics2014;@ellisTyphlopid2019]. The diversification is reported to have originated from a single colonisation event by a non-arid adapted ancestor from Asia between 19-39 mya [@vidalBlindsnake2010;@marinTracing2013], before the onset of aridification over the past 20 million years [@byrneBirth2008]. 

However, the study did not account for alternative biogeographic areas (e.g. WWF Ecoregions) and models, or uncertainty in phylogenetic relationships and divergence times. Despite representing a substantial portion of snake biodiversity, there has been few investigations into the macroevolutionary dynamics of blind snakes. The monophyly of the genus makes _Anilios_ well suited to test hypotheses about how lineages diversified with dramatic ecological changes because they are geographically restricted to the continent and can be found in almost every biome. 

<!-- Environmental correlates of diversification -->
<!-- agamid lizards [@harmonTempo2003], passerine birds [@ricklefsGlobal2006], amphibians [@rollandContribution2019], rosid plants [@sunRecent2020]).  -->
A variety of ecological correlates have been associated with diversification processes of a clade. Among others, studies on the influence of climate (e.g. temperature, precipitation, and latitude) [@ricklefsGlobal2006;@mayhewBiodiversity2012;@condamineMacroevolutionary2013;@cantalapiedraDietary2014;@condamineClimate2019;@rollandContribution2019;@sunRecent2020], elevation [@esquerreHow2019;@boschmanMountain2021], biological interactions [@ramirezAsynchronous2011;@druryEstimating2016;@brennanPhylogenomicsMonitorLizards2021], CO<sub>2</sub> [@kergoatOpposite2018], microhabitats [@bars-closelDiversification2017] and aridification [@raboskyExceptional2007;@pepperGenetic2011;@deepakAridification2018;@catulloAridification2014] have been documented. In Australia, shifts in diversification rates owing to changes in temperature and aridity has been reported in skinks [@raboskyExceptional2007], geckos [@fujitaDiversification2010;@pepperGenetic2011;@pepperPalaeoclimate2011;@oliverIndependent2014;@oliverContrasting2014], frogs [@catulloAridification2014;@vidal-garciaConvergent2015], saltbush [@kadereitMolecular2010], *Acacia* [@millerComparative2013], insects [@owenHow2017;@toussaintUnveiling2015], and other plants (reviewed in @crispHow2013).

There are several links between temperature and diversification rates of reptiles. First, warm temperature may increase resource availability, which in turn increase ecological niche in general [@clarkeClimate2006]. Second, spatial and temporal heterogeneity in temperature may lead to climatic niche divergence, such as local adaptations in different geographic areas [@pepperGeodiversity2013], and change in activity time of prey/predators [@santosTemperature2011;@lawsonLatitudinal2014]. Third, change in climate (e.g. glaciation and aridification) may cause fragmentation of continuous habitats leading to isolated populations, which may lead to divergence [@oliverRocks2019]. Fourth, ectotherms are subjected to higher metabolic rates in warm climates (i.e. Kleiber's law), which may increase mutation rates and ultimately greater speciation rates [@allenKinetic2006;@makarievaMean2008]. Finally, aridification of Australia may offer new ecological niche for species to exploit, which have been shown to result in deep divergences in many arid-adapted [@byrneBirth2008;@pepperGenetic2011;@marinTracing2013]. 

The influence of temperature on diversification rates has been empirically tested with birth-death models [@condamineMacroevolutionary2013;@kergoatOpposite2018;@condamineClimate2019;@rollandContribution2019], these temperature data are reconstructed (interpolated) based on conversion of concentration of oxygen isotopes [@condamineMacroevolutionary2013]. However, the relationship between diversification rates and aridity are attributed to some proxy such as the the timing of aridification or global temperature rather than quantifying the extent of aridity itself [@dornStrange2014;@deepakAridification2018;@catulloAridification2014;@fujitaDiversification2010;@pepperGenetic2011;@pepperPalaeoclimate2011;@oliverIndependent2014;@oliverContrasting2014;@owenHow2017;@crispHow2013;@toussaintUnveiling2015]. As such, no study has been able to quantitatively assess aridity as a continuous variable in environment-dependent models and compare it to other ecological processes. Recent development in geographic-based methods such as `gen3sis` modeling engine [@hagenGen3sis2021] enables us to simulate climatic, geological, and ecological factors that are spatially and temporally specific to certain areas. In practice, these outputs from `gen3sis` provide paleotemperature and aridity data specific to Australia in the time span of interest, between the 45 MYA to present. These datasets will allow us to empirically test the influence of localised ecological processes on diversification rates over time. 

In this study, we use the Australian radiation of blindsnakes as a system to investigate the diversification of morphologically conserved squamates by testing hypotheses about the processes that promote speciation.

<!--Why people should care-->
- Applying the cutting edge localised geographic-based paleoenvironment reconstruction, 

We 
- Estimate the ancestral biome of Australian blindsnakes based on extant taxa.
- Compare fit_env models between localised versus globalised temperature on the diversification of Australian blindsnake radiation, 
- Assess model fit between paleo-aridity and paleo-climate on diversification rates, and 
- Are there differences in the rates of speciation between arid and non-arid zones?

#  Materials and methods

## Phylogenetic hypothesis and taxon sampling
<!-- Based on our phylogenetic tree -->
We employ the recent time-calibrated phylogenomic tree of Australian blindsnake with outgroups. This tree includes 38 operational taxonomic units (OTUs) representing  __taxa representing __ described species of _Anilios_. However, because our phylogenetic data suggest some cryptic species, we adjust our taxon sampling fraction to XX/XX species. The genetically distinct OTUs that are not fully recognised species include 3 '_Anilios grypus_' and 2 '_Anilios ligatus_'. 

## Biogeography 

We used the R package `BioGeoBEARS` 1.1.1 [@matzkeBioGeoBEARS2013] to estimate the ancestral biome of blindsnakes in Australia. BioGeoBEARS provides a flexible likelihood-based framework that models how the branches in the phylogeny (representing extant taxa and their ancestors) have shifted among a set of discrete biomes over time. These shifts in biomes can occur within a single branch (anagenetic events including processes like dispersal, extinction, or switching to adjacent biomes) or at the time of branching (cladogenetic events involving sympatric, vicariant and founder-event speciation). The fundamental inputs for the software include a time-calibrated phylogeny, information about the biome/range each species are found, and a historical biogeographic model. We divided Australo-Papua into nine discrete biomes and assigned each Australian species to specific biome(s) based on their occurence records as described above. Biome classification were based on the World Wildlife Fund's (WWF) Terrestrial Ecoregions of the World dataset [@olsonTerrestrial2001] and followed the nomenclature of previous Austrlia-specific studies [@cardilloPhylogeny2017;@pavon-vazquezComprehensive2021]. 
The phylogeny and distribution information were then applied to widely used historical biogeographical methods including Dispersal-Vicariance Analysis (DIVA; [@ronquistDispersalVicariance1997]), and likelihood implementations of Dispersal-Extinction-Cladogenesis (DEC; [@reeMaximum2008]) or DECLIKE, and BayArea [@landisBayesian2013] or BAYAREALIKE [@matzkeBioGeoBEARS2013]. We included two additional free parameters. The first parameter *j* assigns per-event weight to founder-events [@matzkeModel2014], and the second parameter _x_ modifies the probability of dispersal based on distance between biomes [@vandamEvaluating2016] (see Appendix-method-biogeobears). We used likelihood ratio tests to determine if the additional free parameter is a justified improvement over the simpler base models, and selected the best overall model based on corrected Akaike Information Criterion (AICc) scores. 

## Reconstruction of Australian paleo climate and environment-dependent diversification analyses
<!-- Consult Alex Skeels -->
Three paleographic digital elevation models (DEMs) providing paleotopography at resolution of 1&deg; x 1&deg; and 5 myr resolution  were obtained from @scoteseAtlas2021, @straumeGlobal2020, [VALDES]. We projected the paleoenvironmental reconstructions into an equal-area Behrmann projection of 110km x 110km following methods described in @hagenGen3sis2021. From the projected data we extracted temperature and aridity values across all grid cells that fall within Sahul by manually drawing a polygon that encompassed the continent from 40 mya to the present-day. Because reptiles are sensitive to extremely low and high temperature, we decided to to summarise the reconstructed paleo temperature data into three different data sets: minimum temperature, maximum temperature, and mean temperature. 

To test whether and how diversification rates correlated with climate and aridity through time, we fitted several time- and environemnt-dependent likelihood diversification models using the `fit_bd` and `fit_env` function in the R package `RPANDA` [@morlonRPANDA2016] to the _Anilios_ phylogeny. Specifically, we fitted different combinations of pure-birth and birth-death models with speciation and/or extinction rates being constant or otherwise varying linearly or exponentially with the following environment variables: 1 mean global paleo temperature, 3 Australian paleo temperature, and 3 Australian paleo aridity. We accounted for missing taxa using the fractioning function (f = XX) in RPANDA. We then assessed which diversification model best fit the data based on AICc scores. 

To visualise the dynamics of the radiation, we use ClaDS [@malietFast2022] implemented in `julia` to estimate speciation rates at the origin of each branch (hencforth branch-specific rates). ClaDS is designed to capture frequent shifts with small effects. At each speciation event, daughter lineages are assigned its own speciation rate sampled from a distribution parameterised on the parental rate [@malietFast2022]. 

## Biogeographic-dependent diversification analyses

We further investigate whether broad patterns of diversification are associated with geographic range. Using the R package `hisse` v. 2.1.9 [@beaulieuDetecting2016], we fitted five biogeographic-dependent speciation-extinction models from the GeoSSE [@goldbergPhylogenetic2011] and GeoHiSSE [@caetanoHidden2018] framework to the tree and geographic range data. 
Since state speciation-extinction (SSE) models only allow three states, we restricted our geographic states to arid, non-arid, and widespread. Consistent with our BioGeoBEARS anaylsis, we classified species that primarily occur in the arid zone as arid (n = 8), species that occur primarily outside of the arid zone as non-arid (n = 24), and those that occupy both the arid and non-arid zone as widespread (n = 5). We assumed uniform taxon sampling for 89% in arid areas, 71% in non-arid areas, and 83% in widespread areas based on number of species in the phylogeny and all currently described species of _Anilios_.
The five models vary by whether they allow diversification rates to be associated with geographic range and/or with unobserved hidden states. Model 1 assumes equal diversification rates regardless of geographic range, model 2 allows geographic range to affect diversification rates (canonical GeoSSE model). Model 3 and 4 are derivation of models 1 and 2 but with unobserved hidden states (see Table GeoHiSSE table). The last model is a complementary model that removes cladogenetic effect entirely. We compared the AIC scores to select the best supported models, then averaged the estimates of diversification parameters and ancestral geographic range estimate across all models according to their AIC weights following @caetanoHidden2018. 

## Statistical analyses
All statistical analyses were run in R version 4.2 (CITE R). We visualised diversification data using `ggplot2`, made map with `QGIS`. 

# Results

## Biogeographic history
We reconstructed the tropical grassland as the ancestral biome of Australian _Anilios_ (\@ref(fig:fig-biogeobears). The preferred biogeographic model is the constrained-dispersal matrix, DEC+j+x. 
Likelihood ratio test (LRT) indicates that founder event (parameter j) and distance matrix (free parameter _x_) in the more complex model provides statistically significant improvement over the simpler DEC, DEC+*j*, and DEC+*x* models (X^2^, DF = 2, D-statistic = 26.7, p-value < 0.001). 
Our result also supports the hypothesis that the ancestor of the Australian clade invaded the continent, but exactly where they lived prior to this invasion is equivocal between the Solomon Islands and New Guinea. 
After arriving to Australia in the late Oligocene - early Miocene (~25 MYA), the ancestral _Anilios_ initially diversified rapidly within less than 1 million years. 
- Our analysis is consistent with @marinTracing2013 conclusion that Australian blindsnakes, like several other Australian squamates, originated from tropical-mesic origins and expanded into the arid zone since the mid-Miocene [@chappleParallel2004;@melvilleEvolutionary2011]. 

## Biome transitions and current distributions
Following the hypothesized colonisation event from a single common ancestor, the diversifcation of _Anilios_ occured very rapidly. Descendents of the earliest split remained in the Tropical Grassland biome but within a million years later another split occured. 
The earliest biome transition was from tropical grassland into Temperate Forest around 25 MYA (95% HPD: 28 – 22 MYA) and the Arid zone 24 MYA (95% HPD: 28 – 21 MYA). 
Subsequent splits show transitions from the arid zone into the Mediterranean biome starting after 15 MYA. 
Current distributions of _Anilios_ are mostly concentrated in the tropical grassland biome closely followed by the arid zone and Mediterranean biome. 
Closely related species tend to share the same biome and occur around the same geographical areas. 

- Range expansion was important for diversification ...
- Topography dyanmics [@brazModelling2021]
- Solomon Island 40 MYA [@hallCenozoic2002]

Recent biogeographic reconstruction recovered India as the origin for superfamily Typhlopoidea (Gerrhopilidae, Xenotyphlopidae and Typhlopidae) [@kleinEvolution2021]. 

## Paleoclimate estimates

Paleoclimate reconstructions of aridity differed slightly between the two methods \@ref(fig:fig-paleoclimate-compare)a. Both aridity index shows increase in aridity post 10 mya, consistent with the onset of aridification of the continent [@byrneBirth2008;@pepperLife2021]. Temperature for Australia differed between the three methods \@ref(fig:fig-paleoclimate-compare)b. Reconstructed temperature values from Australia were higher than global temperature and stayed relatively constant compared to global temperature that shows slight decrease. 

## Diversification
The best fitting model among constant, time-dependent, temperature-dependent, and aridity dependent models is 
This model indicate that X associate with speciation (alpha = ) and .....
 
Fit with Australia specific mean temperature receives better support than model with global temperature and exponential dependency of speciation rate with time. 

We found no statistical difference between diversification rates between arid and non-arid adapted lineages. 
Rates of diversification between arid and non-arid, and widespread. 




\newpage

# Discussion

## Localised climate model outperforms global models

## Ancestral biome reconstruction
- Our result — that the ancestor of Anilios occupied tropical grasslands – is consistent with the result presented in @marinTracing2013. 
- We found transitions from XXX to XXX to be common as have been shown / as opposed to other Australian squamates (CITATIONS)
- Arid zone was not arid until ~20 Million Years ago. Ancestors in deeper time probably occupied temperate or tropical areas that subsequently became the arid zone we know see today. 
- The Australo-Papuan clade seems to have stayed in Australia for the most part but we were not able to include New Guinean endemic _Anilios erycinus_, the only recognised species of _Anilios_ not found in Australia. Therefore, we cannot make any conclusions about transitions between Australia and New Guinea even though at least three species (_A. torresianus_, _A. leucoproctus_, ...) are found in Northern Australia Torres Strait Islands, and southern parts of New Guinea.

## Drivers of diversifcation 
- We expected the biogeographic models with jump dispersal to be the best fitting model because _Anilios_ are believed to have non-Australian origins [@marinTracing2013;@vidalBlindsnake2010;@oliverPhylogenetic2017].
- Blind snakes are thought to not move very much and so formation of geological barriers during aridification may accelerate speciation between 20 – 10 MYA. 
- The fact that DEC+J+x model was preferred over DEC+J may be due to poor dispersal ability as closely related species occupy the same or nearby areas. Events like floods are likely to have faciliated dispersals but in short distances. 
Discuss species that are far apart...
- High rates of diversification observed in concert with changing climate may be explained by allopatric speciation. 
- Speciation could even be sympatric but no evidence to support this. 

## Niche evolution

# Conclusion


Account for heterogeneity of the arid zone [@pepperLife2021].

# Tables

Table BioGeoBEARS. BioGeoBEARS model estimates and model AICc scores. Lower AICc scores indicate model is preferred. + Log likelihood
Table RPANDA. Diversification model fit estimates and AICc score of model fit

# Figure legends

Fig. Ancestral biome: Stochastic map of Ancestral biome from Australo-Papuan blind snakes
Fig. Paleoclimate: Reconstructed paleoclimate values from different data sets. Mean temperature data. Aridity index based on precipitation threshold. Abbreviations Pli = Paleocene and Ple = Pleistocene. 
Fig. Speciation rates + phylogeny over time + lineages through time

\newpage 

**References**
<!-- # References -->